---
title: "Getting started with jrSiCKLSNMF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with jrSiCKLSNMF}
  %\VignetteEngine{knitr::knitr}
  %\usepackage[UTF-8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Loading jrSiCKLSNMF and Creating a SickleJr Object
Here, we will provide a walkthrough of jrSiCKLSNMF on simulated data. First, install and load the jrSiCKLSNMF package:

```{r loadjrSiCKLSNMF}
library(jrSiCKLSNMF)
```

For this walkthrough, we will be working with simulated data object \textsc{SimData}. These data were generated from GSE130399 using the packages SparSim and simATAC. The details for the simulations can be found in our paper. \textsc{SimData} has already been QC'd; however, when working with real data, you should QC your data and select features that appear in at least 10 cells for both views. After loading \textsc{jrSiCKLSNMF} into R, you should have access to \textsc{SimData}.

Now, we create a SickleJr object. This object will hold all of the results and intermediary steps for jrSiCKLSNMF. We will also create a vector containing the cell types for the simulated data for use in the next section.

```{r createSickleJr}
data("SimData")
DataMatrices<-SimData$Xmatrices
cell_type<-SimData$cell_type
SimSickleJr<-createSickleJr(SimData$Xmatrices)
rm(DataMatrices,SimData)
```

# Adding Metadata

Next, we will add metadata in the form of the true cell identity to the SickleJr object.

```{r addmetadata}
SimSickleJr<-AddSickleJrMetaData(SimSickleJr,cell_type,"true_cell_type")
rm(cell_type)
```

You can add any metadata to the SickleJr object. Make sure to set the variable \textsc{metadata} to the variable containing the metadata you wish to add and the metadataname to a string containing what you wish the metadata to be called. Here, we call the metadata "true_cell_type." 

# Generation of Graph Laplacians for Graph Regularization

Next, we build our KNN graphs and extract the graph Laplacian to use for graph regularization. Note that for large matrices, this can take some time. Please set a seed if you want to have the exact same graph Laplacians every time.

```{r graphlaplacian}
SimSickleJr<-BuildKNNGraphLaplacians(SimSickleJr)
```

# Normalization, Initialization of W and H, and Setting Lambda Values

Next, we will normalize our count matrices using median library size normalization.

```{r Normalize}
SimSickleJr<-NormalizeCountMatrix(SimSickleJr)
```

We then initialize our W and H matrices below.

```{r initializeWH}
SimSickleJr<-GenerateWmatricesandHmatrix(SimSickleJr)
```

For values $\lambda_{W^v}$ and our value $\lambda_H$, when we constrain the rows of $\textbf{H}$ such that the L2 Norm of each row is equal to 1 (the default), we suggest setting $\lambda_{\textbf W^{RNA}}=3$, $\lambda_{\textbf W^{ATAC}}=15$, and $\lambda_{\textbf H}=0$. When not using this constraint, we suggest using $\lambda_{\textbf W^{RNA}}=10$, $\lambda_{W^{ATAC}}=50$, and $\lambda_{H}=500$. 

```{r setlambdas}
SimSickleJr<-setLambdas(SimSickleJr,list(3,15))
```

# Running jrSiCKLSNMF

Finally, we can run jrSicKLSNMF. By default, we constrain the rows of $\textbf{H}$ such that the L2 Norm of each row is equal to 1. Please note that we store $\textbf{H}$ as $\textbf{H}^T$. Please note that, for real data, you should use enough rounds to reach convergence of $10^{-6}$ for the L2 Norm regularized data and $10^{-7}$ for the sparsity data.

```{r runningjrsickls}
SimSickleJr<-RunjrSiCKLSNMF(SimSickleJr,rounds=20000)
```
# Post-hoc Analyses

After this, we can perform cell clustering. In version 1.0, users can use k-means or spectral clustering. Here, we demonstrate k-means

```{r clustering}
SimSickleJr<-ClusterSickleJr(SimSickleJr,numclusts=3)
```

Finally, we can calculate and then plot the UMAP of our SickleJr. Note that if you would like to plot the UMAP of the compressed $\mathbf{W}^v\mathbf{H}$ matrix, please enter the number corresponding to the view you wish to see. For the plots, you can either color based off of identified cluster or based off of metadata.

```{r UMAPplots}
SimSickleJr<-CalculateUMAPSickleJr(SimSickleJr)
#Plotting based off of cluster
plotSickleJrUMAP(SimSickleJr,title="K-means clusters")
#Plotting based off of true cell type metadata
plotSickleJrUMAP(SimSickleJr,colorbymetadata="true_cell_type",title="True Cell Types",legendname="True Cell Types")
```

We can also visualize data in the RNA view and the ATAC view. 

```{r UMAPplotsWH}
SimSickleJr<-CalculateUMAPSickleJr(SimSickleJr,view=1)
plotSickleJrUMAP(SimSickleJr,title="K-means clusters: RNA view",
                 umap.view="W1H")
plotSickleJrUMAP(SimSickleJr,colorbymetadata="true_cell_type",
                 title="True Cell Type: RNA view",legendname="True Cell Types",
                 umap.view="W1H")
SimSickleJr<-CalculateUMAPSickleJr(SimSickleJr,view=2)
plotSickleJrUMAP(SimSickleJr,title="K-means clusters: ATAC view",umap.view="W2H")
plotSickleJrUMAP(SimSickleJr,colorbymetadata="true_cell_type",
                 title="True Cell Type:ATAC view",legendname="True Cell Types",
                 umap.view="W2H")
```
